Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-12 bounds.tex
Refactor introduction.

%\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
\section{Dual-tree algorithms}

%\begin{itemize}
%\item What are pairwise statistical problems? Examples?
%
%A surprising number of machine learning algorithms have bottlenecks which, if
%implemented naively, depend on the calculation of potentially large numbers of
%distances.  A common example is nearest neighbor search: if search
%for the nearest neighbor of a single query point in a reference set of size $N$,
%simple linear scan costs $O(N)$ time.  For many decades, spatial data structures
%have been employed to accelerate this type of computation, beginning with the
%quad tree \cite{finkel1974quad}; building trees on the reference set often
%results in expected runtimes of $O(\log N)$ per query point.
%
A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

%\item What are dual-tree algorithms? Brief explanation.
%
%More recently, Gray and Moore \cite{nbody} observed that problems often have
%large query {\em sets}, not just a few query points, and introduced dual-tree
%algorithms.  One example is all-nearest-neighbor search: find the nearest
%neighbor of every point in the reference set.  Simple implementations scale
%poorly, but dual-tree algorithms can obtain orders-of-magnitude speedup for a
%multitude of problems.
%
%Following this, numerous dual-tree algorithms have been developed to solve a
%wide variety of problems; these include kernel density estimation
%\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
%approximate nearest neighbor search \cite{ram2009rank}, mean shift
%\cite{wang2007fast}, minimum spanning tree calculation
%\cite{march2010euclidean}, $n$-point correlation function estimation
%\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
%smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
%
Gray and Moore \cite{nbody} generalized the fast-multipole-method from
computational physics to obtain {\em dual-tree algorithms}. These are extremely
useful when there are large query {\em sets}, not just a few query points.
%\item What (pairwise statistical) problems have they been applied to?
%
Following this, numerous dual-tree algorithms have been developed to solve a
wide variety of problems; these include kernel density estimation
\cite{gray2003nonparametric}, fast max-kernel search \cite{curtin2013fast},
approximate nearest neighbor search \cite{ram2009rank}, mean shift
\cite{wang2007fast}, minimum spanning tree calculation
\cite{march2010euclidean}, $n$-point correlation function estimation
\cite{march2012fast}, kernel summation \cite{lee2006faster}, particle
smoothing \cite{klaas2006fast}, variational inference \cite{amizadeh2012variational} and range search \cite{nbody}.
%
%\item Existing analyses of dual-tree algorithms
Some of these algorithms are derived using the cover tree
\cite{langford2006}, a data structure with beneficial theoretical qualities.
Dual-tree all nearest-neighbor search and approximate kernel density estimation with cover trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}. Other problems have similar worst-case guarantees \cite{curtin2014dual, march2013multi}.
% Is there a better reference for the npcf runtime bound?

Gray and Moore \cite{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Dual-tree
algorithms can be easily understood through the recent framework of Curtin
et~al. \cite{curtin2013tree}: two trees (a query tree and a reference tree) are
traversed by a {\it pruning dual-tree traversal}.  This traversal visits
combinations of nodes from the trees in some sequence (each combination consisting of a query node and a
reference node), calling a problem-specific \texttt{Score()} function to
determine if the node combination can be pruned.  If not, then a
problem-specific \texttt{BaseCase()} function is called for each combination of
points held in the query node and reference node.  This has significant
similarity to the more common single-tree branch-and-bound algorithms, except
that the algroithm must recurse into child nodes of {\it both} the query tree
and reference tree.

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
minimum spanning tree calcul

%\item General tree/problem independent dual-tree algorithms (maybe)
%\item What is lacking -- general dual-tree analysis (since the algorithms are so similar in structure)
%A recent framework for understanding and unifying dual-tree algorithms
%observes that all dual-tree algorithms are particularly similar, and generalizes
%them \cite{curtin2013tree}.  A
%nice benefit of this viewpoint is that it allows quick and easy development of
%new dual-tree algorithms.  In this paper, we make the same observation about
%runtime proofs for cover tree dual-tree algorithms, and show that
%runtime analysis of generalized cover tree dual-tree algorithms is
%possible.  Thus, the contributions of this paper are:
%
%\begin{itemize}
%  \item A general runtime bound for dual-tree algorithms that use the cover
%tree.
%
%  \item \mbox{A nearest neighbor search runtime bound,
%slightly improving on existing linear-time results.}
%
%  \item An approximate kernel density estimation runtime bound,
%improving on existing linear-time results.
%
%  \item Novel runtime analysis of dual-tree range count and range
%search.
%\end{itemize}
Curtin et.~al. \cite{curtin2013tree} observe that all dual-tree
algorithms are similar and propose a unified framework which allows the
development of dual-tree algorithms in a problem independent (and tree
independent) manner. In this paper, we make the same observation about runtime
proofs for cover tree dual-tree algorithms, and show that runtime analysis
of a generalized dual-tree algorithm is possible. The specific contributions of
this paper are:
\begin{itemize}
  \item A general runtime bound for dual-tree algorithms that use the cover
tree.
  \item An improved linear runtime guarantee for all nearest-neighbor search.
  \item An improved linear runtime guarantee for approximate kernel density
estimation.
  \item The first linear runtime guarantee for dual-tree range count and range
search.
\end{itemize}

%The following section will detail theory prerequisites, and then introduce the
%general runtime bound.

%\item This paper's contribution:
%  \begin{itemize}
%  \item Problem independent dual-tree algorithm analysis
%  \item Examples of final problem dependent runtime bounds for NNS and KDE which match existing results (from LTAPS) -- implying nothing is lost yet for doing problem independent analysis
%  \item Novel runtime analyses of dual-tree range count (or range search).
%  \end{itemize}
%\item This paper's structure (to be written last)
%\end{itemize}

\section{Preliminaries}

@@

\section{Tree Imbalance}

@@

\section{General runtime bound}

@@

\section{Nearest neighbor search}

@@

\section{Approximate kernel density estimation}

@@

\section{Range Search and Range Count}

>>>>

\section{Dual-tree algorithms}

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

Gray and Moore \cite{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Instead of
building a tree on the reference set and searching with each query point
separately, Gray and Moore suggest also building a query tree and traversing
both the query and reference trees simultaneously (a {\it dual-tree traversal},
from which the class of algorithms takes its name).

Dual-tree algorithms can be easily understood through the recent framework of
Curtin et~al. \cite{curtin2013tree}: two trees (a query tree and a reference
tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
visits combinations of nodes from the trees in some sequence (each combination
consisting of a query node and a reference node), calling a problem-specific
\texttt{Score()} function to determine if the node combination can be pruned.
If not, then a problem-specific \texttt{BaseCase()} function is called for each
combination of points held in the query node and reference node.  This has
significant similarity to the more common single-tree branch-and-bound
algorithms, except that the algroithm must recurse into child nodes of {\it
both} the query tree and reference tree.

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
minimum spanning tree calculation \cite{march2010euclidean}, $n$-point
correlation function estimation \cite{march2012fast}, max-kernel search
\cite{curtin2013fast}, particle smoothing \cite{klaas2006fast}, variational
inference \cite{amizadeh2012variational}, and range search \cite{nbody}, to name
a few.

Some of these algorithms are derived using the cover tree \cite{langford2006}, a
data structure with beneficial theoretical qualities.  Dual-tree all
nearest-neighbor search and approximate kernel density estimation with cover
trees have $O(N)$ runtime guarantees for $O(N)$ queries \cite{ram2009}; minimum
spanning tree calculation scales as $O(N \log N)$ \cite{march2010euclidean}.
Other problems have similar worst-case guarantees \cite{curtin2014dual,
march2013multi}.

In this work we combine the generalization of Curtin et~al.
\cite{curtin2013tree} with the theoretical results of Beygelzimer et~al.
\cite{langford2006} (and others) in order to develop a worst-case runtime bound
for any dual-tree algorithm when the cover tree is used.

Section \ref{sec:preliminaries} lays out the required background, notation, and
introduces the cover tree and its associated theoretical properties.  Readers
familiar with the cover tree literature and dual-tree algorithms (especially
\cite{curtin2013tree}) may find that section to be review.  Following that, we
introduce an intuitive measure of cover tree imbalance in Section
\ref{sec:imbalance}.  This measure of imbalance is then used to prove the main
result of the paper in Section \ref{sec:bound}, which is a worst-case runtime
bound for generalized dual-tree algorithms.  We apply this result to three
specific problems: nearest neighbor search (Section \ref{sec:nns}), approximate
kernel density estimation (Section \ref{sec:akde}), and range search / range
count (Section \ref{sec:rs}), showing linear runtime bounds for each of those
algorithms.  Each of these bounds is an improvement on the state-of-the-art, and
in the case of range search, is the first such bound.  We then investigate the
empirical performance of cover trees for these algorithms, with a focus on
understanding the dataset-dependent constants that our bounds use.

\section{Preliminaries}
\label{sec:preliminaries}

@@

\section{Tree Imbalance}
\label{sec:imbalance}

@@

\section{General runtime bound}
\label{sec:bound}

@@

\section{Nearest neighbor search}
\label{sec:nns}

@@

\section{Approximate kernel density estimation}
\label{sec:akde}

@@

\section{Range Search and Range Count}
\label{sec:rs}
